Performance benefits of Intel Optane DC storage for the parallel processing of large neuroimaging data

### 2019-09-18 report/report.tex

\newcommand{\valerie}[1]{\color{blue}\textbf{Note from Valerie}:
      #1 \color{black}}

@@

\title{Performance benefits of Intel Optane DC storage for the processing of large neuroimaging data}

@@

\begin{abstract}
    Open-access neuroimaging datasets have reached up to $\SI{}{\peta\byte}$ of data, and
    continue to grow. The ability to leverage the entirety of these datasets are
    limited to a restricted number of labs with both the time and infrastructure
    to process the data. Whereas BigData engines have significantly reduced
    application performance penalties with respect to data movement, their applied
    strategies are not necessarily practical within neuroimaging workflows where
    intermediary made need to be materialized to shared storage for post-processing
    analysis. In this paper we evaluate the performance advantage brought by Intel
    Optane DC persistent memory storage for the processing of large neuroimaging 
    datasets using the two available configurations modes: Memory Mode and App Direct Mode.
    We employ a synthetic algorithm on the $\SI{76}{\gibi\byte}$ \bigbrain
    and a real neuroimaging application on the CoRR dataset using 25 and 96 parallel
    processes in both cases. It was found that performance of all available storage
    devices (e.g. tmpfs, local disk and Isilon) is somewhat deteriorated by memory mode
    when compared to App Direct Mode. In contrast, Optane in App Direct Mode was
    found to be slower than local disk, when the load on the bus was high due to sheer
    amount of concurrent processes and short task durations. It is believed that
    Optane would likely be useful in the accessing of typically larger-than-memory
    data in a limited-parallelism manner, such as with visualization applications.
    Further experiments with larger datasets remain to be performed.

@@

\section{Introduction}
Neuroimaging open-data initiatives have led to extensively large repositories of
publicly available data. Such initiatives include the \bigbrain~\cite{BigBrain}, 
a one-of-a-kind $\SI{603}{\gibi\byte}$
histological image of a 65 year-old healthy human brain
at 20$\mu$m resolution; the UK Biobank~\cite{ukbiobank}, a repository expected to
contain approximately $\SI{0.2}{\peta\byte}$ of data (including various magnetic
resonance (MR) imaging modalities) from 500,000 individuals living in the UK;
the Human Connectome Project~\cite{HCP}, a repository containing MR scans from
1200 healthy adults, which is expected to exceed $\SI{2}{\peta\byte}$ in size; 
and the Consortium for Reliability and Reproducibility (CoRR)~\cite{corr}, an
initiative which aggregates MR data from various laboratories around the world, 
of which 32 are currently available and make up about $\SI{937}{\giga\byte}$ of
data in total.

Due to storage limitations, only subsets of the neuroimaging repositories 
can be processed in a typical research laboratory. Moreover, as these datasets 
are extremely large and are only increasing in size, they are typically stored 
in higher-capacity, slower storage devices. In such conditions, large-scale 
studies in neuroimaging remain limited to labs with access to adequate 
infrastructure and are limited by data transfer times.

@@

Whereas popular Big Data engines, such as MapReduce~\cite{mapreduce} and Apache Spark~\cite{spark}, have
incorporated software solutions to limit data transfers (e.g. in-memory computing,
data locality, and lazy evaluation), hardware has also adapted to the growing datasets.
One such improvement is the concept of placing flash storage directly on the Dual
In-line Memory Module (DIMM), thereby reducing the latency of accessing data on 
flash storage devices. While the latency of these devices is improved, the bandwidth
remains the same. However, as noted by \cite{nvdimms}, a severe performance degradation
can be experienced by having memory traffic and I/O placed on the same bus.

@@

@@

@@

For the purposes of these experiments, we use the inequality introduced in
\cite{paperinmem}. That is,
\begin{equation}                                                                 
\frac{D}{C} \leq \frac{\delta}{\gamma} \label{eq:page-cache-inequality}                                                                                                                                  
\end{equation} 
where:
\begin{itemize}
    \item $D$ is the total amount of data written by the application
    \item $C$ is the total execution time of the application
    \item $\delta$ is the disk bandwidth
    \item $\gamma$ is the max number of concurrent processes on a node
\end{itemize}

Using this inequality, we anticipate that whenever the application is within the limits,
the storage selection will make little difference as storage devices with writeback enabled
will be able to leverage the memory cache. It is expected that Optane in App Direct Mode
will be most useful when this inequality is not satisfied. As Memory Mode will inevitably
increase the amount of data that can be cached by the storage, it is expected that the overall
performance of all storage devices, even when the inequality is not satisfied.

@@

Due to the size and uniqueness of the \bigbrain, standardized processing pipelines
have yet to be developed for it. In order to quantify the effects of data location
on such a dataset, we have implemented a naive synthetic application that takes 
the image, split into blocks, and increments all the voxels within each block, in parallel (Figure~\ref{alg:incrementation}).
This application
enables us to read and write to different storage devices and ensure that written
data could not have been previously cache
in-memory, by ensuring that read and written data are not the same. This application
will be parallelized using both GNU Parallel and Apache Spark 2.4.3 and implemented in Python 3.6. While the GNU Parallel application
is currently limited to 1 incrementation step, the Apache Spark implementation allows us to increase
the number of incrementation steps by a variable amount, thereby increasing the amount of intermediate
data generated.

@@

\subsubsection{ BIDS App Example}

@@

As shown in Figure~\ref{fig:makespan-25cpus}, despite being under the inequality, the
performance still varies according to selected storage device. As expected, Optane performs
slightly better than local disk and significantly better than Isilon. Tmpfs outperforms all
devices, with only slightly better performance than Optane and local disk, in both App Direct Mode and Memory Mode.
There is no apparent difference between the two modes on the given devices.

When observing the total task breakdown (Figure~\ref{fig:stacked-25cpus}), it is
apparent that the differences between the makespans arise from difference in incrementation
time between all devices. While tmpfs and local disk have comparable write times, Optane
has a more significant write time difference. Whereas, Isilon has more significant read and
write time compared to all other devices, it's incrementation time is comparable to that
of local disk. Further inspection of each storage device's Gantt chart (see Figure~\ref{fig:gantt25})
show that increased duration of tasks stagger the execution of each individual tasks. This, in turn,
results in better performance towards the end of the execution, where tasks are significantly more staggered.
Only tmpfs appears to be able to execute each task efficiently with little staggering.

\subsubsection{96 Processes}
\begin{figure}
    \includegraphics[width=\columnwidth]{../experiments/bigbrain-incrementation/figures/makespan-96cpus-0delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{Makespan of \bigbrain Incrementation using 96 processes on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:makespan-96cpus}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{../experiments/bigbrain-incrementation/figures/stacked-96cpus-0delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{total read/increment/write breakdown of \bigbrain incrementation using 96 processes on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:stacked-96cpus}
\end{figure}

@@

As with 25 process, the main cause of performance variability appears to be the incrementation
time (Figure~\ref{fig:stacked-96cpus}. Local disk running in App Direct Mode has an average
incrementation time that is nearly 15~x faster than in Memory Mode. The incrementation time
for local disk is, on average, 6~x faster than on Optane.

@@

\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{../experiments/bids-app-example/25cores-sum.pdf}
    \caption{25 processes}\label{fig:bb25}
\end{subfigure}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{../experiments/bids-app-example/96cores-sum.pdf}
    \caption{96 processes}\label{fig:bb96}
\end{subfigure}
\caption{I/O and CPU breakdown of BIDS App Example 25 and 96 processes on all storage devices. 3 repetitions were performed}\label{fig:bbd}
\end{figure*}

Unlike the BigBrain incrementation at 25 Processes, the makespan of the BIDS App Example
executing on Optane is longer than local disk (Figure~\ref{fig:bm25}). However, similarly
to the BigBrain Incrementation, tmpfs has a slight lead over local disk and Isilon performs the
worst. The same pattern can be found with 96 process (Figure~\ref{fig:bm96}). When comparing
the executions of 25 and 96 processes, the performance improved by a similar factor on all
storage devices.

@@

While Memory Mode provides a cost-effective solution to increasing the total memory
size, it is slower than App Direct Mode for embarrassingly parallel applications such
as \bigbrain incrementation. This is likely due to the Page Cache being composed of a
combination of both DRAM and Optane in Memory Mode. Since the data produced by the application should be
able to fit within DRAM, writing to a slower bandwidth device such as Optane inevitably
deteriorates performance. As App Direct separates usage of DRAM and Optane, devices only utilize DRAM in this mode.
The only device that performed better in Memory Mode was 
Isilon. This may be explained by a larger Page Cache in Memory Mode, enabling Isilon to
store more data in memory prior to initiating synchronous writes to Isilon.

I/O-less task duration also seem to be affected by the execution mode, with those tasks
also performing significantly better in App Direct. A potential explanation for this may be 
that all memory accesses to Optane are slower due to the devices bandwidth, meaning that
tasks that do not perform I/O, but access data loaded into memory are slower in Memory Mode.
This can be seen in Figure~\ref{fig:gantt96} where for local disk, incrementation in App Direct
mode is significantly shorter. Only when the memory bus is overwhelmed does incrementation take
longer for App Direct, however, it is nothing compared to the average duration of the incrementation
tasks in Memory Mode.

>>>>

\newcommand{\valerie}[1]{\color{blue}\textbf{Note from Valerie}:
      #1 \color{black}}
\newcommand{\tristan}[1]{\color{orange}\textbf{Note from Tristan}:
      #1 \color{black}}

@@

\title{Performance benefits of Intel Optane DC storage for the parallel processing of large neuroimaging data}

@@

\begin{abstract}
    Open-access neuroimaging datasets have reached up to $\SI{}{\peta\byte}$ of data, and
    continue to grow. The ability to leverage the entirety of these datasets are
    limited to a restricted number of labs with both the capacity and infrastructure
    to process the data. Whereas BigData engines have significantly reduced
    application performance penalties with respect to data movement, their applied
    strategies are not necessarily practical within neuroimaging workflows where
    intermediary results may need to be materialized to shared storage for post-processing
    analysis. In this paper we evaluate the performance advantage brought by Intel
    Optane DC \tristan{expan acronym?} persistent memory storage for the processing of large neuroimaging 
    datasets using the two available configurations modes: Memory Mode and App Direct Mode.
    We employ a synthetic algorithm on the $\SI{76}{\gibi\byte}$ \bigbrain
    and a real neuroimaging application on the CoRR \tristan{expand acronym} dataset using 25 and 96 parallel
    processes in both cases. Our results show that performance of all available storage
    devices (e.g. tmpfs, local disk and Isilon) is somewhat deteriorated by memory mode
    when compared to App Direct Mode. In contrast, Optane in App Direct Mode was
    found to be slower than local disk, when the load on the bus was high due to sheer
    amount of concurrent processes and short task durations. It is believed that
    Optane would likely be useful in the accessing of typically larger-than-memory
    data in a limited-parallelism manner, such as with visualization applications.
    Further experiments with larger datasets remain to be performed.

@@

\section{Introduction}
Neuroimaging open-data initiatives have led to extensively large repositories of
publicly available data. Such initiatives include the \bigbrain~\cite{BigBrain}, 
a one-of-a-kind $\SI{603}{\gibi\byte}$
histological image of a 65 year-old healthy human brain
at 20$\mu$m resolution; the UK Biobank~\cite{ukbiobank}, a repository expected to
contain approximately $\SI{0.2}{\peta\byte}$ of data (including various magnetic
resonance (MR) imaging modalities) from 500,000 individuals living in the UK;
the Human Connectome Project~\cite{HCP}, a repository containing MR scans from
1,200 healthy adults, which is expected to exceed $\SI{2}{\peta\byte}$ in size; 
and the Consortium for Reliability and Reproducibility (CoRR)~\cite{corr}, an
initiative which aggregates MR data from various centres around the world, 
of which 32 are currently available and make up about $\SI{937}{\giga\byte}$ of
data in total.

Due to storage limitations, only subsets of such neuroimaging repositories 
can be processed in a typical research laboratory. Moreover, as these datasets 
are extremely large and are only increasing in size, they are typically stored 
in higher-capacity, slower storage devices \tristan{such as?}. In such conditions, large-scale 
studies in neuroimaging remain limited to labs with access to adequate 
infrastructure and are limited by data transfer times.

@@

Whereas popular Big Data engines, such as MapReduce~\cite{mapreduce} and Apache Spark~\cite{spark}, have
incorporated software solutions to limit data transfers (e.g. in-memory computing,
data locality, and lazy evaluation), hardware has also adapted to the growing datasets.
One such improvement is the concept of placing flash storage directly on the Dual
In-line Memory Module (DIMM), thereby reducing the latency of accessing data on 
flash storage devices. While the latency of these devices is improved, the bandwidth
remains the same. However, as noted by \cite{nvdimms}, a severe performance degradation
can be experienced by having memory traffic and I/O placed on the same bus \tristan{references are
missing or the bib file hasn't been updated}.

@@

\tristan{awesome intro Valerie, great context, right level of detail, nice 
parallel between software and hardware solutions for big data, super well informed.}

@@

\tristan{I guess tmpfs, Isilon and local disks should be described here.}

@@

For the purposes of these experiments, we use the inequality introduced in
\cite{paperinmem} to characterize data-intensive applications. That is,
\begin{equation}                                                                 
\frac{D}{C} \leq \frac{\delta}{\gamma} \label{eq:page-cache-inequality}                                                                                                                                  
\end{equation} 
where:
\begin{itemize}
    \item $D$ is the total amount of data written by the application
    \item $C$ is the total execution time of the application
    \item $\delta$ is the disk bandwidth
    \item $\gamma$ is the max number of concurrent processes on a node
\end{itemize}

Using this inequality, we anticipate that whenever the application is within the limits,
the storage selection will make little difference as storage devices with writeback enabled \tristan{This should be explained 
before, in methods (infrastructure) and possibly in intro.}
will be able to leverage the memory cache. It is expected that Optane in App Direct Mode
will be most useful when this inequality is not satisfied \tristan{why?}. As Memory Mode will inevitably
increase the amount of data that can be cached by the storage, it is expected that the overall
performance of all storage devices \tristan{will what?}, even when the inequality is not satisfied.

@@

Due to the size and uniqueness of the \bigbrain, standardized processing pipelines
have yet to be developed for it. In order to quantify the effects of data location
on such a dataset, we have implemented a naive synthetic application that takes 
the image, split into blocks, and increments all the voxels within each block, in parallel (Algorithm~\ref{alg:incrementation}).
This application
enables us to read and write to different storage devices and ensure that written
data could not have been previously cache
in-memory, by ensuring that read and written data are not the same. This application
will be parallelized using both GNU Parallel and Apache Spark 2.4.3 and implemented in Python 3.6 \tristan{In a subsequent 
version you could give more details about implementations, including containerization, etc}. While the GNU Parallel application
is currently limited to 1 incrementation step, the Apache Spark implementation allows us to increase
the number of incrementation steps by a variable amount, thereby increasing the amount of intermediate
data generated.

@@

\subsubsection{BIDS App Example}

@@

As shown in Figure~\ref{fig:makespan-25cpus}, despite being under the inequality, the
performance still varies according to selected storage device. As expected, Optane performs
slightly better than local disk and significantly better than Isilon. Tmpfs outperforms all
devices, with only slightly better performance than Optane and local disk, in both App Direct Mode and Memory Mode.
There is no apparent difference between the two modes on the given devices.
\tristan{You should explain why Optane is not reported in memory mode.}

When observing the total task breakdown (Figure~\ref{fig:stacked-25cpus}), it is
apparent that the differences between the makespans arise from difference in incrementation
time between all devices except Isilon. In addition, while tmpfs and local disk have comparable write times, Optane
has a more significant write time difference. Whereas, Isilon has more significant read and
write time compared to all other devices, it's incrementation time is comparable to that
of local disk. Further inspection of each storage device's Gantt chart (see Figure~\ref{fig:gantt25})
show that increased duration of tasks stagger the execution of each individual tasks. This, in turn,
results in better performance towards the end of the execution, where tasks are significantly more staggered.
Only tmpfs appears to be able to execute each task efficiently with little staggering.
\tristan{In the methods, you should say how time was measured.}
\tristan{Is Isilon in writeback? If so, why is write time so large?}
\subsubsection{96 Processes}
\begin{figure}
    \includegraphics[width=\columnwidth]{../experiments/bigbrain-incrementation/figures/makespan-96cpus-0delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{Makespan of \bigbrain Incrementation using 96 processes on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:makespan-96cpus}
\end{figure}
\begin{figure}
    \includegraphics[width=\columnwidth]{../experiments/bigbrain-incrementation/figures/stacked-96cpus-0delay-1it.pdf}
    \captionsetup{width=\columnwidth}
    \caption{total read/increment/write breakdown of \bigbrain incrementation using 96 processes on 125 blocks on the 40~$\mu$m image given different storage configurations. 5 repetitions were performed.}\label{fig:stacked-96cpus}
\end{figure}

@@

As with 25 process, the main cause of performance variability appears to be the incrementation
time (Figure~\ref{fig:stacked-96cpus}). Local disk running in App Direct Mode has an average
incrementation time that is nearly 15~x faster than in Memory Mode. The incrementation time
for local disk is, on average, 6~x faster than on Optane.

@@

\begin{figure*}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{../experiments/bids-app-example/25cores-sum.pdf}
    \caption{25 processes}\label{fig:bb25}
\end{subfigure}
    \begin{subfigure}{\columnwidth}
        \centering
    \includegraphics[width=\columnwidth]{../experiments/bids-app-example/96cores-sum.pdf}
    \caption{96 processes}\label{fig:bb96}
\end{subfigure}
\caption{I/O and CPU breakdown of BIDS App Example 25 and 96 processes on all storage devices. 3 repetitions were performed}\label{fig:bbd}
\end{figure*}
\tristan{I/O time might include sleep time if the process is throttled, we should explain how I/O and CPU were measured.}
Unlike the BigBrain incrementation at 25 Processes, the makespan of the BIDS App Example
executing on Optane is longer than local disk (Figure~\ref{fig:bm25}). However, similarly
to the BigBrain Incrementation, tmpfs has a slight lead over local disk and Isilon performs the
worst. The same pattern can be found with 96 process (Figure~\ref{fig:bm96}). When comparing
the executions of 25 and 96 processes, the performance improved by a similar factor on all
storage devices.

@@

While Memory Mode provides a cost-effective solution to increasing the total memory
size, it is slower than App Direct Mode for embarrassingly parallel applications such
as \bigbrain incrementation. This is likely due to the Page Cache being composed of a
combination of both DRAM and Optane in Memory Mode. Since the data produced by the application should be
able to fit within DRAM, writing to a slower bandwidth device such as Optane inevitably
deteriorates performance. As App Direct separates usage of DRAM and Optane, devices only utilize DRAM in this mode.
The only device that performed better in Memory Mode was 
Isilon \tristan{But the difference looks very small}. This may be explained by a larger Page Cache in Memory Mode, enabling Isilon to
store more data in memory prior to initiating synchronous writes to Isilon.

I/O-less task duration also seem to be affected by the execution mode, with those tasks
also performing significantly better in App Direct. A potential explanation for this may be 
that all memory accesses to Optane are slower due to the devices bandwidth, meaning that
tasks that do not perform I/O, but access data loaded into memory are slower in Memory Mode.
This can be seen in Figure~\ref{fig:gantt96} where for local disk, incrementation in App Direct
mode is significantly shorter. Only when the memory bus is overwhelmed does incrementation take
longer for App Direct, however, it is nothing compared to the average duration of the incrementation
tasks in Memory Mode. \tristan{how about throttling?}
